Stack Safety as a Security Property

### 2020-07-01 main.tex
Comments and polishing on 2

Before we define stack safety, we need to first set the stage by
describing our machine model. Our definition of stack safety is
generic and this sections discusses an abstract interface to our
machine model and some kind of static or dynamic, software- or
hardware-enforced security monitor. At the same time, for
concreteness, we will build up our running example using a
conventional machine architecture with a standard (RISC-V) processor
and later extend it with a dynamic policy mechanism.

@@

The basic building blocks of the machine are {\em values} and {\em
  addresses}. Both are drawn from some set of all possible words
$\WORDS$. We will denote words with $\word$.
%
Our machine states are comprised of {\em components}, that can be word
(denoting memory addresses) or registers (drawn from some set $\REGS$).
%
    \[\reg \in \REGS\]
    \[\component \in \COMPONENTS = \WORDS + \REGS \]
%
The registers must include two special purpose registers: the program
counter {\PCname} and the stack pointer \SP.
%
Formally, a machine state is just a map from components to values:
%
\[\mach \in \MACHS = \COMPONENTS \rightarrow \WORDS\]

Our stack safety property will be defined in terms of {\em
  observations}, which may be silent (written $\tau$) or a word of
data:\leo{Needs more exposition}
%
\[\obs \in \OBSS = \tau ~ | ~ \word \in \WORDS\]
%
These observations are a byproduct of the step function of the
machine, which takes a machine state and produces a pair of a
machine state and an observation.
%
\(\mach \stepstoobs{\obs} \mach' : \MACHS \rightarrow \MACHS \times \OBSS \).

@@

Mapping a conventional architecture to this abstract interface is
straightforward. To build up our running example, we are going to use
a standard 64-bit RISC machine, just like the one we are using in our
experiments later on \leo{Mention FORVIS?}. In this setting, $\WORDS$
is the set of 64-bit integers, $\REGS$ is the set $\{r_0, \cdots,
r_{31}\} \cup {\PCname}$, and $\SP = r_2$. \leo{Mention We use the
  standard encoding of instructions to words?}

The first interesting choice is that of observations, as there are
multiple reasonable options. At one extreme, we could choose to define
observations as the value written at any component. However, that
would pose problems later on when we treat some components as ``secret''
when they correspond to a caller's private state. \leo{On the other extreme,
we could pick a particular component and treat it as the ``output'' of
our system. }

@@

A policy is a general model of an enforcement mechanism, consisting of
a set of policy states \(\pol \in \POLS\), and a policy step function
\((\mach, \pol) \stepstopol \pol' : \MACHS \times \POLS
\rightharpoonup \POLS\).

We write $\MPS$ for the set of all pairs of machine states and policy
states.
%
\[\MPS = \MACHS \times \POLS\]
%
We then lift this policy step function to operate on such pairs,
by combining it with the regular step function for machine states.

@@

\leo{Notation: Should $\stepstoobs{}$ use harpoon for the lifted step?}

\leo{TODO: Forward pointer to policies right?}

@@

A trace over some type $A$ is a potentially infinite, nonempty
sequence of $A$s. We will mostly be concerned with sequences of
machines states, usually denoted $\machT$, and with sequences of
triples of (machine states, policy states, and observations), usually
denoted $\MPT$.
\rb{The way the definitions are set up, one may expect MPO, but\ldots}
To continue, we will require four generic operations
on traces: a ``trace-of'' operation to construct the potentially
infinite traced induced by a step relation
\rb{not an op on traces per se?};
a ``split'', to break a
trace into two components based on a predicate; a ``join'', the
reverse of split; and a ``prefix'', as a shorthand to only keep the
first component of a split.

The ``trace-of'' operator, written \(\hookrightarrow\), coinductively
relates a machine state with the trace of machine states paired with
observations produced by repeated application of step:
%
\[\hookrightarrow : \machT \rightarrow \MOTRACE\]%
%
\judgmenttwo{\(\mach_0 \stepstoobs{\obs} \mach_1\)}{\(\mach_1 \hookrightarrow \machT\)}
            {\(\mach_0 \hookrightarrow (\mach_0,\obs) \machT\)}%
%
\noindent
Since the step function is total, if \(\mach \hookrightarrow \machT\),
then \(\machT\) must be infinite.

\leo{TODO: Add an example of a trace once I figure out the notations}

Similarly, we define \(\hookrightarrow_P\) to relate a machine-policy
state pair to the trace of triples of machine states, policy states,
and observations induced by by \(\stepstoobs{}_P\):
%
\[\hookrightarrow_P : \MACHS \times \POLS \rightarrow \MPOTRACE \]
%
\judgment{\(\neg \exists \pol_1 . \pol_0 \rightharpoonup \pol_1\)}
         {\(\mpstate{\mach_0}{\pol_0} \hookrightarrow \mpostate{\mach_0}{\pol_0}{\tau}\)}
%
\judgmentthree{\(\mach_0 \stepstoobs{\obs} \mach_1\)}
              {\(\pol_0 \rightharpoonup \pol_1\)}
              {\(\mpstate{\mach_1}{\pol_1} \hookrightarrow \machT\)}
              {\(\mpstate{\mach_0}{\pol_0} \hookrightarrow_P \mpostate{\mach_0}{\pol_0}{\obs} \machT\)}
We will also use \(\pi_m\) to project the trace of machine states
out of a trace, \(\pi_p\) to project the trace of policy states, and
\(\pi_o\) to project the observations.

% SNA: fixed
%\leo{PUT is completely ambiguous as a choice of abbreviation. Just ``Pre''
%  seems better}
The {\it prefUpTo} operation \(\PUT(\Trace,f)\) takes a trace \(T\)
and a predicate on elements \(f\) and gives the prefix of \(T\)
ending with the first element on which \(f\) holds, coinductively:

\judgment{\(f\ \traceelem\)}
         {\(\PUT(f,\traceelem\Trace) = \traceelem\)}
%
\judgment{}
         {\(\PUT(f,\traceelem) = \traceelem\)}
%
\judgmenttwo{\(\neg f\ \traceelem\)}{\(\PUT(f,\Trace) = \Trace'\)}
            {\(\PUT(f,\traceelem\Trace) = \traceelem\Trace'\)}
%
Note that if the resulting trace is a strict prefix of the parameter, then the
predicate must hold on its final element. Conversely, if the predicate never
holds, \(\PUT\) is the identity function.

>>>>

To set the stage, we first
describe our machine model. Our definition of stack safety is
generic and this section discusses an abstract interface to our
machine model and some kind of static or dynamic, software- or
hardware-enforced security monitor.  For
concreteness, our running examples will use a
conventional machine architecture and a standard (RISC-V) instruction set;
in~\Cref{sec:micropolicies} we will
extend this model it with a dynamic policy enforcement mechanism based
on \bcp{NickPaper,AndMaybeSomeOthers}.

@@

The basic building blocks of the machine are {\em values} and {\em
  addresses}. Both are drawn from some set of {\em words}
$\WORDS$, ranged over by $\word$.
%
Our machine states are composed of {\em components} $k$, addressed either by
words (memory addresses) or by by register names ($r$) drawn from some set
$\REGS$.
%
    \[\component \in \COMPONENTS = \WORDS + \REGS \]
%
The register names are assumed to include two special purpose registers: the
program counter {\PCname} and the stack pointer \SP.

A machine state is a map from components to values:
%
\[\mach \in \MACHS = \COMPONENTS \rightarrow \WORDS\]

The step function of the
machine takes a machine state and produces a pair of a
machine state and an {\em observation}, which may be either
silent (written $\tau$) or a word of data that the machine communicates to
the outside world (e.g., by writing to a DMA address):\leo{Needs more
  exposition}\bcp{Like that?}
%
\[\obs \in \OBSS = \{\tau\} \uplus \WORDS\]
\[\mach \stepstoobs{\obs} \mach' : \MACHS \rightarrow \MACHS \times \OBSS \]
%

@@

Mapping a conventional architecture to this abstract interface is
straightforward. To build up our running example, we are going to use
a standard 64-bit RISC machine, just like the one we are using in our
experiments later on \leo{Mention FORVIS?}. In this setting, $\WORDS$
is the set of 64-bit integers, $\REGS$ is the set $\{r_0, \cdots,
r_{31}\} \cup \{\PCname\}$, and $\SP = r_2$.
% \leo{Mention We use the standard encoding of instructions to words?}
% BCP: Nah.

The first interesting choice is that of observations, as there are
multiple reasonable options. At one extreme, we could choose to define
observations as the value written at any component. However, that
would pose problems later on when we treat some components as ``secret''
when they correspond to a caller's private state. \leo{On the other extreme,
we could pick a particular component and treat it as the ``output'' of
our system. }  \bcp{But in this paragraph we are describing the particular
choise that we make for our running example, no?  But we certainly want to
discuss this fully someplace!}

@@

\bcp{More explanation needed, plus examples.}A policy is a general model of an enforcement mechanism, consisting of
a set of policy states \(\pol \in \POLS\), and a policy step function
\((\mach, \pol) \stepstopol \pol' : \MACHS \times \POLS
\rightharpoonup \POLS\).

We write $\MPS$ for the set of all pairs of machine states and policy
states.
%
\[\MPS = \MACHS \times \POLS\]
%
We lift the policy step function to operate on such pairs
by combining it with the regular step function for machine states.

@@

\leo{Notation: Should $\stepstoobs{}$ use harpoon for the lifted
  step?}\bcp{Yes.}

\leo{TODO: Forward pointer to policies right?}\bcp{Yes.  But also we can say
in more detail that the intention is to capture any kind of dynamic
enforcement mechanism and give as many names and citations as we can.}

@@

A trace over a type $A$ is a nonempty, finite or infinite
sequence of $A$s. We will mostly be concerned with sequences of
machines states, usually\bcp{not always?} denoted $\machT$, and with sequences of
triples of (machine states, policy states, and observations), usually
denoted $\MPT$.
\rb{The way the definitions are set up, one may expect MPO, but\ldots}
To continue\bcp{??}, we will require four generic operations
on traces: a ``trace-of'' operation to construct the
trace induced by a step relation and a machine state
\rb{not an op on traces per se?}\bcp{+1};
a ``split'', to break a
trace into two components based on a predicate; a ``join'', the
reverse of split; and a ``prefix'', as a shorthand to only keep the
first component of a split.  \bcp{Not sure it's worth enumerating them
  rather than just getting down to the actual definitions below.}

The ``trace-of'' operator, written \(\hookrightarrow\), coinductively
relates a machine state with the trace of machine states paired with
observations produced by repeated application of step:  \bcp{above, we
  showed operators applied to metavariables, not just the bare operator
  symbol.}
%
\[\hookrightarrow : \machT \rightarrow \MOTRACE\]%
%
\judgmenttwo{\(\mach_0 \stepstoobs{\obs} \mach_1\)}{\(\mach_1 \hookrightarrow \machT\)}
            {\(\mach_0 \hookrightarrow (\mach_0,\obs) \machT\)}%
%
\noindent
Since the step function is total, if \(\mach \hookrightarrow \machT\),
then \(\machT\) must be infinite.

\leo{TODO: Add an example of a trace once I figure out the notations}
\bcp{Let's also try to clean up the spacing a little -- many of these
  sequences of rules can be typeset on fewer lines}

Similarly, we define \(\hookrightarrow_P\) to relate a machine-policy
state pair to the trace of triples of machine states, policy states,
and observations induced by by \(\stepstoobs{}_P\): \bcp{Observe that this
  time the trace can be either finite or infinite.}
%
\[\hookrightarrow_P : \MACHS \times \POLS \rightarrow \MPOTRACE \]
%
\judgment{\(\neg \exists \pol_1 . \pol_0 \rightharpoonup \pol_1\)}
         {\(\mpstate{\mach_0}{\pol_0} \hookrightarrow \mpostate{\mach_0}{\pol_0}{\tau}\)}
%
\judgmentthree{\(\mach_0 \stepstoobs{\obs} \mach_1\)}
              {\(\pol_0 \rightharpoonup \pol_1\)}
              {\(\mpstate{\mach_1}{\pol_1} \hookrightarrow \machT\)}
              {\(\mpstate{\mach_0}{\pol_0} \hookrightarrow_P \mpostate{\mach_0}{\pol_0}{\obs} \machT\)}
We will also use \(\pi_m\) to project the trace of machine states
out of a trace, \(\pi_p\) to project the trace of policy states, and
\(\pi_o\) to project the observations.  \bcp{I'm getting a little lost in
  all these.  E.g., do these traces include the final state of a finite
  execution?  If so, what observation is it paired with?}

% SNA: fixed
%\leo{PUT is completely ambiguous as a choice of abbreviation. Just ``Pre''
%  seems better}
The {\it prefUpTo} operation \(\PUT(\Trace,f)\) \bcp{Why does it have two
  different abbreviated names?  And wouldn't ``While'' be more mnemonic?}
takes a trace \(T\) 
and a predicate on elements \(f\) and gives the prefix of \(T\)
ending with the first element on which \(f\) holds, coinductively:

\judgment{\(f\ \traceelem\)}
         {\(\PUT(f,\traceelem\Trace) = \traceelem\)}
%
\judgment{}
         {\(\PUT(f,\traceelem) = \traceelem\)}
%
\judgmenttwo{\(\neg f\ \traceelem\)}{\(\PUT(f,\Trace) = \Trace'\)}
            {\(\PUT(f,\traceelem\Trace) = \traceelem\Trace'\)}
%
Note that if the resulting trace is a strict prefix of the parameter, then the
predicate must hold on its final element. Conversely\bcp{why conversely?}, if the predicate never
holds, \(\PUT\) is the identity function.
